---
title: "Final Project Proposal"
author: "Makenzie Kerneckel"
date: last-modified
institute: "University of Pennsylvannia" 
toc: true 

format: 
  html:
    self-contained: true
    
editor: source
---


## Research question

- For my final project, I want to examine the relationship between political events, specifically close election outcomes between rival political parties in Mexico, and homicide rates at the municipal level. My research question aims to contribute to Melissa Dell's findings in her 2015 paper "Trafficking Networks and the Mexican Drug War". The homicide rate data I will use comes from a github that aggregates data from the National Institute of Statistics and Geography (INEGI). The homicide data spans from 2000 to 2021. The election returns data comes from a github made by Eric Magar who compiled official vote returns from Mexico. Unlike Dell, I do not have access to drug-related homicides. 
  - Homicide github: <https://github.com/thegargiulian/mexico-homicide-rates>
  - Election Return github: <https://github.com/emagar/elecRetrns?tab=readme-ov-file#orgdc3d940>

- For the experiment, I want to conduct a close election regression discontinuity design. The close election variable refers to muncipal elections that were closely contested between the conservative PAN party and other political parties in Mexico. These close elections will serve as a quasi-experimental setting to isolate the variation in crime policy and political leadership changes, allowing for the examination of the impact of PAN party victories on homicide outcomes in municipalities. By comparing municipalities where the PAN party won by a small margin to those where it lost by a small margin, I am trying to isolate the causal effect of PAN victories on homicides. 

- To expand on Dell's design, I can go down two routes. The first is to examine whether municipalities where PAN has historically had a strong power base, lost, then regained power show an decrease in homicide rates. A second route would be to explore another outcome variable: political violence/assassinations. Looking at this second outcome, the question would what is the causal effect of PAN victories on political violence. I could use political violence data from José Ramón Enríquez's "Democracy under Assault: Electoral Reform and Political Violence". 
    - Dell Paper: <https://www.aeaweb.org/articles?id=10.1257/aer.20121637>
    - Enríquez Paper: <https://scholar.harvard.edu/enriquez/publications/democracy-under-assault-electoral-reform-and-political-violence>
    
## Dataset

- Data Cleaning: 

``` {r}
 
library(foreign)
library(tidyverse)
library(magrittr)

onUser<-function(x){
  user<-Sys.info()["user"]
  onD<-grepl(x,user,ignore.case=TRUE)
  return(onD)
}

if(onUser("kenziekmac")){
  figures<-"~/Users/kenziekmac/Dropbox/Mac/Documents/GitHub/psci3200mk/Figures"
  tables<-"~/Users/kenziekmac/Dropbox/Mac/Documents/GitHub/psci3200mk/Tables"
}



election_returns <- read.csv("Data/aymu1970-on.coalSplit.csv")

homicide_rates <- readr::read_delim("Data/mexico-muni-month-homicide-rates-2000-2021.csv", delim = "|")

# have leading zeros so all municipal level identifiers are 5 digits

head(unique(homicide_rates$ent_mun))

# does not have leading zeros 

head(unique(election_returns$inegi))

# let's add leading zeros to the election_returns data 

# Function to add leading zeros

add_leading_zeros <- function(x) {
  ifelse(nchar(x) == 4, paste0("0", x), as.character(x))
}

# Apply the function to the column

election_returns$inegi <- as.numeric(election_returns$inegi)

election_returns$inegi <- sapply(election_returns$inegi, add_leading_zeros) 


election_returns %<>% rename( "ent_mun" = "inegi")
  
# Check for municipal codes present in df1 but not in df2

# these are missing election returns/ newly created municipalities 

records_only_in_df1 <- anti_join(election_returns, homicide_rates, by = "ent_mun")

# drop these municipalities since they don't have corresponding homicide rates 

ent_mun_drops_er <- records_only_in_df1$ent_mun

# 34,634 rows at start 

election_returns %<>% filter(!is.na(ent_mun)) %>%
                      filter(! ent_mun %in% ent_mun_drops_er)

# 34,624 after, dropped 10 rows 


# Check for municipal codes present in df2 but not in df1

# these are missing homicide rates

records_only_in_df2 <- anti_join(homicide_rates, election_returns, by = "ent_mun")

# drop these municipalities since they don't have corresponding election returns 

ent_mun_drops_hr <- records_only_in_df2$ent_mun

# 651,288 rows at start 

homicide_rates %<>% filter(!is.na(ent_mun)) %>%
  filter(! ent_mun %in% ent_mun_drops_hr)

# 650,232 after, dropped 1056 rows

# pick rows we need

election_returns %<>% select(-c(1,3,5,8)) 

# break election date into month and year so we can merge on these variables to match monthly homicide rates

election_returns$date<- as.character(election_returns$date)
  
election_returns %<>% mutate( date = as.Date(election_returns$date, format = "%Y%m%d"),
                              year = format(date, "%Y"),
                              month = format(date, "%m"))

# pick rows we need, again

election_returns %<>% select(-c(1,4))

# filter years bc homicide data starts at 2000

election_returns %<>% filter(year >= 2000)

# add leading zeros to homicide month col.


add_leading_zeros_2 <- function(x) {
  ifelse(nchar(x) == 1, paste0("0", x), as.character(x))
}

# Apply the function to the column


homicide_rates$month <- as.numeric(homicide_rates$month)

homicide_rates$month <- sapply(homicide_rates$month, add_leading_zeros_2) 


# merge 

election_returns$year <- as.numeric(election_returns$year )

test <- full_join(homicide_rates,election_returns, by = c("ent_mun", "month", "year"))
 
```


- Final Dataset:
    
``` {r}

# take a small random sample of the merged data 

# years without elections show NAs for election columns

set.seed(1244)
sample_frac(test, 0.00001)

```
    
 - I will need to calculate a vote share variable. I will do this in the following steps:
    - Determine the total votes cast for the PAN party (PAN_votes) and the total votes cast for all other parties (other_votes). 
    - Calculate the margin of victory or defeat for each election as the absolute difference between PAN_votes and other_votes.
    - Define a threshold for what constitutes a "closely contested" election.
    - Filter the elections where the margin of victory or defeat is within the defined threshold.




