---
title: "Assignment 1"
author: "Makenzie Kerneckel"

institute: "University of Pennsylvania"
date: last-modified
toc: true

format: 
  html:
    self-contained: true

editor: source
---


# Part 1: Read-in data and prepare for analysis

```{r }
#| warning: false
#| message: false

library(ggplot2)
library(readr)
library(ggdag)
library(tidyverse)
library(gt)
library(modelsummary)

# read-in data
# dat = read_csv(here::here("workshops/aau_survey/clean_endline_did.csv" )) %>%
dat = read_csv("https://raw.githubusercontent.com/jrspringman/psci3200-globaldev/main/workshops/aau_survey/clean_endline_did.csv" ) %>%
    # clean home region variable
  mutate(q8_baseline = ifelse(q8_baseline == "Southern Nations, Nationalities, and Peoples Region", "SNNPR", q8_baseline), 
         q8_baseline = str_remove(q8_baseline, " Region"))
# create color palette for plotting
palette = MetBrewer::met.brewer(name = "Cross")

```


## Requirement 1 

For all variables after user_language, please rename the column with a descriptive name that better conveys their meaning. Column names should never contain spaces and should be as easy to type as possible. Do this for both their baseline and endline values, making sure to indicate which columns are baseline measures and which are endline measures in the names you assign.

```{R}

# rename and select responseid ,gender,q26, q27, q 17 columns 

dat <- dat %>%
 select(1,"gender_base" = "q3_baseline",
         "civil_base" = "q26_civ_baseline",
         "politics_base" = "q26_politics_baseline",
         "public_base" = "q26_public_baseline",
         "poloffice_base" =  "q27_1_baseline",
         "ngo_base" = "q27_3_baseline",
         "positive_indv_base" = "q17_3_baseline",
         "youth_engage_base" = "q17_1_baseline",
         "positive_youth_base" = "q17_2_baseline",
         "civil_end" = "q26_civ",
         "politics_end" = "q26_politics",
         "public_end" = "q26_public",
         "poloffice_end" = "q27_1",
         "ngo_end" = "q27_3",
         "positive_indv_end" = "q17_3",
         "youth_engage_end" = "q17_1",
         "positive_youth_end" = "q17_2",
         )


``` 


# Part 2: Create Index Measures

## Requirement 2 (10%)

First, in your own words, explain the concept of an additive index and an averaged z-score, including how they are calculated, when you should use them, and when you cannot use them. What are the benefits of each approach?

- An additive index adds individual values or variables into a single score. It should be used when you have many variables measuring a single concept and these variables are measured on a common scale. It should also be used when you are interested in measuring the accumulation of something. This can be for outcome, treatment, and covariate variables. The main benefits of the additive index are that it simplifies graphs and hypotheses by condensing multiple variables into one metric. This makes it easier to compare individuals or groups. 
  - To calculate an additive scale you simply sum across columns (index = col_1 + col_2...).
- A z-score tells you the number of standard deviations a value is from the mean of the given distribution of your data. A z-score is a standardized score. An averaged z-score combines each standardized score for your values by averaging them. Average z-scores should be used when variables have different scales or when variables cannot be summed in a dataset. The benefit of average z-scores is that it allows you to interpret a variable through standard deviations.  
  - To calculate an averaged z-score, you first standardize the scores for each variable (subtract the mean and divide by the standard deviation), then you average these standardized scores.

## Requirement 3 (20%)

Create an additive index for the baseline and endline measures of the “Future plans for a career in public sector or civil society” variables. This should correspond to seperate counts of the number of future plans that each individual has at baseline and endline.

```{R}

# convert future plans columns into binary values of 1 and 0 using as.logical 

eth_svy <- dat %>% 
  mutate(across(c(3:7), ~+as.logical(.x))) %>%
  mutate(across(c(11:15), ~+as.logical(.x))) 

# create an additive index variable called add_plans for baseline and endline 

# in this case na values represent no anwser so we drop them from the sum

eth_svy <- eth_svy %>%
  # baseline sums across all future plans questions at baseline 
  mutate(add_plans_end =  rowSums(across(11:15), na.rm = T)) %>%
  # baseline sums across all future plans questions at baseline 
  mutate(add_plans_base =  rowSums(across(3:7), na.rm = T))


```

Create an averaged z-scores for the baseline and endline values of the “Future plans for a career in public sector or civil society” and “Feelings of political efficacy” variables.


```{R}

#| warning: false



# create a list of baseline column names 

# select all the columns with base in the name

bcols = grep("base", names(eth_svy), value = T)

# drop gender_base

bcols = bcols[-c(1)]

eth_svy[, paste0(bcols, "_st")] = eth_svy[, bcols]

bcols = paste0(bcols,"_st")

# do the procedure above but for endline

ecols = grep("base", names(eth_svy), value = T, invert = T) 

# invert function gives everything without base in the name

# drop response id column 

ecols = ecols[-c(1)]

eth_svy[, paste0(ecols, "_st")] = eth_svy[, ecols]

ecols = paste0(ecols,"_st")


# Modified z-score function
z_score <- function(x) {
  mean_val <- mean(x, na.rm = TRUE)
  sd_val <- sd(x, na.rm = TRUE)
  (x - mean_val) / sd_val
}

# Calculate z-scores using a loop
for (i in c(bcols, ecols)) {
  eth_svy[[i]] <- z_score(eth_svy[[i]])
}
# create average z-score columns 

eth_svy <- eth_svy %>% 
  rowwise() %>% 
  # avg. z score for future plans question endline 
  mutate( z_future_end = mean(c_across(all_of(ecols[1:5])), na.rm = TRUE)) %>% 
  # avg. z score for future plans question baseline 
  mutate( z_future_base = mean(c_across(all_of(bcols[1:5])), na.rm = TRUE)) %>%
  # avg. z score for political efficacy question endline 
  mutate( z_pol_end = mean(c_across(all_of(ecols[6:8])), na.rm = TRUE)) %>% 
  # avg. z score for political efficacy question baseline 
  mutate( z_pol_base = mean(c_across(all_of(bcols[6:8])), na.rm = TRUE)) %>%
  ungroup()

```


## Requirement 4 (20%)

To make sure that these scores look as you’d expect, create a ggplot visualizing the distribution of the z-scores at baseline and endline. You should have 4 figures: one corresponding to each z-score at baseline and endline. In words, describe whether the figures tell us anything about changes over time.


```{R}

# create four histograms to show endline and baseline distribution on average z scores for future plans and political question


ggplot(eth_svy , aes(x = z_future_base )) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), 
                 binwidth= 0.5, fill = palette[9]) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent of respondents", x = "Z-Score", 
       title = "Public Sector or Civil Society Career Plans (Baseline)") +
  # Add a vertical line at the mean value of the data with a red dashed line
  geom_vline(xintercept = mean(eth_svy$z_future_base, na.rm = T), col = 'red', lwd = 1, linetype = 'dashed')



ggplot(eth_svy , aes(x = z_future_end )) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), 
                 binwidth= 0.5, fill = palette[9]) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent of respondents", x = "Z-Score", 
      title = "Public Sector or Civil Society Career Plans (Endline)") +
  # Add a vertical line at the mean value of the data with a red dashed line
  geom_vline(xintercept = mean(eth_svy$z_future_end, na.rm = T), col = 'red', lwd = 1, linetype = 'dashed')




ggplot(eth_svy , aes(x = z_pol_base )) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), 
                 binwidth=0.5, fill = palette[9]) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent of respondents", x = "Z-Score", 
       title = "Feelings of Political Efficacy (Baseline)") +
     # Add a vertical line at the mean value of the data with a red dashed line
  geom_vline(xintercept = mean(eth_svy$z_pol_base, na.rm = T), col = 'red', lwd = 1, linetype = 'dashed')



 
ggplot(eth_svy , aes(x = z_pol_end )) + 
  geom_histogram(aes(y = after_stat(count / sum(count))), 
                 binwidth=0.5, fill = palette[9]) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent of respondents", x = "Z-Score", 
      title = "Feelings of Political Efficacy (Endline)") +
   # Add a vertical line at the mean value of the data with a red dashed line
  geom_vline(xintercept = mean(eth_svy$z_pol_end, na.rm = T), col = 'red', lwd = 1, linetype = 'dashed')
      




```


In words, describe whether the figures tell us anything about changes over time.

- Given that this survey data contains two waves, the baseline graphs represent student responses collected in May-June whereas the endline graphs represent student responses collected in October-November of 2022. The data is standardized so you can't see obvious changes in the mean or standard deviation over time. You are able see changes in skew and peaks in standardized data but in this case the distributions of z-scores does not change over time from baseline to endline for the future plans or political efficacy variables. 

- If anything, the future plans variable has a right-skewed distribution of z-scores which indicates that the data is not normally distributed and is characterized by a tail extending to the right of the peak of the distribution. In this context the mean of the future plans z-scores is slightly greater than 0, indicating that the average value of the future plans variable is shifted towards the higher end of the scale.




# Part 3: Estimating models

Now, let’s estimate some models to assess the relationship between the two index measures. Before we get started, subset your data to include only response_id, q3_baseline (which you should have renamed), and the baseline and endline measures for each z-score. You should end up with 6 variables in your dataframe.

```{r}

# select 6 variables needed, create new dataset for modeling 

eth_model <- eth_svy %>%
             select( response_id, 
                     gender_base, 
                     z_future_end, 
                     z_future_base,
                     z_pol_end,
                     z_pol_base)

```


## Requirement 5 (15%)


Using baseline values only, estimate a model regressing your “Future plans” index on your “Feelings of political efficacy” index. 



```{r}

 
pol_model <- list()
pol_model[['Bivariate']] = lm(z_future_base ~ z_pol_base, eth_model)

modelsummary(
  pol_model,
  estimate  = "{estimate}{stars} ({std.error})",
             statistic = NULL,
  gof_omit = 'IC|RMSE|Log|F|R2$|Std.')


```

Substantively, how should we interpret the relationship described in the data? What does this tell us about the world? What assumptions would we need in order to interpret the relationship as causal?


- The regression coefficient for the "Feelings of political efficacy" index (z_pol_base) shows a increase by 0.083 units in the "Future plans in Public Sector or Civil Society" index (z_future_end) for a one-unit increase (or an increase of one standard deviation) in "Feelings of political efficacy" index, holding all other variables constant. In other words, students who report higher levels of political efficacy tend to have, on average, slightly more future plans in the public sector or civil society compared to those with lower levels of political efficacy, assuming all other factors remain unchanged.

- We need two main assumptions to interpret the relationship as casual. First, we assume there is no reverse causality or that future plans does not cause changes in political efficacy. Second, we assume that there are no unobserved variables (confounders) that simultaneously affect both political efficacy and future plans. Unobserved confounding occurs when there are variables that influence both the independent and dependent variables but are not included in the analysis. This can lead to biased estimates of the relationship between political efficacy and future plans. In this case, we need to assume that those who believe in political efficacy must be identical in all possible ways to those who do not. 


## Requirement 6 (15%)

For your baseline and endline values of the “Feelings of political efficacy” index, convert this index to a binary indicator taking a value of 1 of the individual has a value greater than or equal to the sample mean and a value of 0 if the individual has a value below the sample mean.

Using baseline values only, estimate the same model, but interact your binary “Feelings of political efficacy” indicator with the gender indicator. Your model should take the following form:


```{r}

eth_model <- eth_model %>%
             mutate(bi_pol_end = ifelse(z_pol_end >= 0, 1,0),
                    bi_pol_base = ifelse(z_pol_base >= 0, 1,0))


bi_pol_model <- list()
bi_pol_model[['Interaction']] = lm(z_future_base ~ bi_pol_base*gender_base, eth_model)

modelsummary(
  bi_pol_model,
  estimate  = "{estimate}{stars} ({std.error})",
             statistic = NULL,
  gof_omit = 'IC|RMSE|Log|F|R2$|Std.')

            
```

In your own words, interpret the meaning of the model and Substantively, how should we interpret the interactive relationship described in the data?


α: The intercept indicates the estimated baseline outcome, which is the average z-score index for future plans, for individuals who are female (gender_baseMale = 0) and have "Feelings of political efficacy" below the sample mean (bi_pol_base = 0), with a value of -0.108. 

β1: This coefficient (0.111) represents the change in the baseline outcome (future plans average z-score index) for individuals whose values on the "Feelings of political efficacy" index change from below the mean (bi_pol_base = 0) to above or equal to the mean (bi_pol_base = 1) at the base time period but only for females (gender_baseMale = 0). However, the p-value (0.066) suggests that this coefficient is not statistically significant at conventional levels.

β2: Gender_baseMale represents the difference in α between gender_baseMale = 0 (females) and gender_baseMale = 1 (males). It shows the change in the intercept (α) when gender changes from its reference category female to male. In simpler terms, it's asking how much the baseline (starting point) of the outcome variable (future plans average z-score index) changes when we switch from one gender category to another. A coefficient of 0.068 for β2 means that, holding all other variables constant (meaning that we are only looking at individuals with "Feelings of political efficacy" below the sample mean (bi_pol_base = 0), males tend to have an outcome value (future plans average z-score index) that is 0.068 units (standard deviations) higher on average compared to females.


β3: Represents the coefficient associated with the interaction between gender and political efficacy. The difference in β1 between Gender=0 and Gender=1 is asking about the change in the coefficient for political efficacy (b1) when gender changes from female to male. Specifically, it's asking how much the effect of political efficacy (b1) on the outcome variable changes when we switch from female to male. A coefficient of -0.011 for β3 indicates the effect of the interaction between gender and political efficacy on the outcome variable, holding all other variables constant. In practical terms, it suggests that the relationship between political efficacy and the future plans variable differs between genders by -0.011 units. Here, the negative coefficient implies that the effect of political efficacy on the outcome variable is weaker (or less positive) for males compared to females, or stronger (or more negative) for females compared to males. However, it's important to note that the interaction effect is not statistically significant according to the p-value (0.077). 

Substantively, the coefficient for the interaction between gender and political efficacy (β3) suggests that the effect of political efficacy on future plans may vary slightly between males and females. However, the non-significant p-value associated with this coefficient (0.077) indicates that this observed difference may not be reliable or meaningful. In other words, the variation in the effect of political efficacy on future plans between genders could be due to chance or random variability in the data rather than a true difference in the relationship.




## Requirement 7 (10%)

Convert the data from ‘wide’ to ‘long’ format, so that each respondent (response_id) has two rows of data; one row is baseline and one row is endline.

Using this new ‘long’ format, estimate the original model, but add unit (response_id) fixed effects. Your model should take the following form:

 
```{r}

# This code is transforming a dataset by pivoting it from wide to long format, specifically for columns related to political efficacy and future plans variables

# It then adds a new column 'time' to indicate whether the data corresponds to the endline (1) or the baseline (0) of the time period

eth_model_long <- eth_model %>%
  pivot_longer(
    cols = c(z_pol_end, z_pol_base, z_future_end, z_future_base), # Pivot columns related to political efficacy and future plans variables
    names_to = c(".value", "time"), # Split column names into '.value' (which will become the new column name) and 'time' (which indicates end or base)
    names_pattern = "z_(\\w+)_(\\w+)", # Use regex pattern to extract variable names and time periods from original column names
    values_to = c("z_pol", "z_future") # Define new column names for the pivoted values
  ) %>%
  mutate(time = ifelse(time == "end", 1, 0)) # Create a new 'time' column where 'end' corresponds to 1 and 'base' corresponds to 0

fixed_model <- list()
fixed_model[['Fixed Effects']] = lm(future ~ pol + factor(response_id), eth_model_long)

modelsummary(
  fixed_model,
  coef_omit = "factor\\(.*",
  estimate  = "{estimate}{stars} ({std.error})",
             statistic = NULL,
  gof_omit = 'IC|RMSE|Log|F|R2$|Std.')

```


In your own words, tell us how the meaning of has changed now that we’ve added fixed effects.

- In the original model, the regression coefficient represents the change in the future plans index for a one-unit increase (or an increase of one standard deviation) in "Feelings of political efficacy" index, holding all other variables constant. In this model, the regression coefficient represents the change in the "Future plans in Public Sector or Civil Society" index variable (future) associated with going from below the mean to above the mean in "Feelings of political efficacy" index variable (pol), while accounting for within-individual variation over time. Specifically, a one-unit increase in feelings of political efficacy (or an increase of one standard deviation) is associated with an average increase of 0.016 units in the future plans in the public sector or civil service average z-score index across respondents, holding constant individual-specific effects that remain constant over time. In simple terms, the first model averages across time and different types of people while this model controls for all variables that don't change over time for respondents.


 